---
title: "HW 4"
author: "Abhijit Dasgupta"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = 'hw_img', fig.ext = 'png')
library(tidyverse)
library(broom)
library(ggpubr)
datadir = '~/Dropbox/teaching/BIOF439/data'
riskfactors <- readRDS(fs::path(datadir,'riskfactors.rds'))
```


In this homework you will generate your own R Markdown file, either using the R Markdown templates in RStudio or from scratch. For each question, you will create R chunks which you will populate with code. The code and the results should be visible in the HTML file that you knit the R Markdown to. 

The header of the R Markdown file will look like this

```
---
title: "Week 4 assignment"
author: "Your name"
output: html_document
---
```

Each question is 10 points.

### Question 1

We will use the dataset `riskfactors` available in the class data link. 

There are two files holding the same data. 
`riskfactors.rds` stores the data in R-native format, which can be read as `riskfactors <- readRDS('data/riskfactors.rds')` (you can update the path based on where you store the file). The `riskfactors.csv` file can be read using `read_csv`. 

```{r, echo=T, eval=F}
riskfactors <- readRDS('data/riskfactors.rds')
```

#### Part a

Draw a summary plot that shows data types and missing data patterns

#### Part b

Draw an upset plot of missing data patterns in this dataset including the top 10 most missing variables.

```{r, echo=TRUE, eval=FALSE}
naniar::gg_miss_upset(riskfactors, _____ = 10)
```


### Question 2

Plot the frequency distribution of marital status in this dataset, as seen below.

```{r, echo=F}
ggplot(riskfactors %>% filter(!is.na(marital)), aes(fct_rev(fct_infreq(marital))))+
  geom_bar() + 
  labs(x = 'Marital status', 
       y = 'Frequency',
       title = 'Frequency distribution of marital status')+
  coord_flip()
```

> Important points for this plot. 
> 
> 1. I do not want a bar for missing values
> 1. The tricky part may be getting the right order. See the forcats cheatsheet for hints
> 1. There is use of `coord_flip`. This should be the last thing you do in the ggplot pipe. 

```{r, echo = TRUE, eval = FALSE}
plot_data <- riskfactors %>% ________

ggplot(plot_data, 
       aes(x = ________(_________(marital)))) + 
  geom_bar() + 
  labs(x = ________, 
       y = ________,
       main = __________) + 
  coord_flip
```


### Question 3

We will use the `USArrests` data set that is inbuilt in R. 

Recreate the following plot in R

```{r, echo=F}
library(ggrepel)
USArrests %>% rownames_to_column('State') %>% 
  ggplot(aes(UrbanPop/100, Murder))+
  geom_text_repel(aes(label = State)) + 
  labs(y = 'Murder arrests (per 100,000)', 
       caption = 'McNeil (1977). Interactive Data Analysis') + 
  scale_x_continuous('Percent of population that is urban',
                     labels = scales::percent) +
  theme_classic()
```

> You will need to add the rownames in the data to become a proper column named `State`  using `USArrests %>% rownames_to_column('State')`


### Question 4

Using the `riskfactors` data, create a panel of plots using any of the three packages introduced in lecture, where the plots will be

1. A histogram of age
1. A frequency bar plot of sex
1. A boxplot of BMI by education level
1. A barplot showing the proportion of each educational level for each marital status.


### Extra credit (10 pts)

Install the **coronavirus** package (`install.packages('coronavirus')`). Once that is installed, run `coronavirus::update_dataset()`. This will update the data to the latest data from JHU, and also ask you to restart
your R session (`Session > Restart R`). 

#### Part a

The `coronavirus` dataset is in the **coronavirus** package and contains data on cases, deaths and recoveries on a daily basis from around the world. We will only use the deaths data, which is obtained by filtering the rows
of the `coronavirus` dataset to contain only rows with `type=="death"`

Extract out data for 6th July, 2020 (`date == "2020-07-06`) and then identify the top 10 countries in terms of
deaths on that date. 

There are a couple of nuances here. For some countries, there is also provincial data, so you have to sum over those rows within country (`%>% group_by(country) %>% summarise(cases = sum(cases, na.rm=T) %>% ungroup()`))

You can use `slice_max` to grab the approriate rows. To pull out the countries, you can then use `%>% pull(country)` and save the result. You should have

```{r, echo=FALSE, message=FALSE}
library(coronavirus)
coronavirus %>% filter(date==max(date), type=='death') %>% group_by(country) %>% summarise(cases = sum(cases, na.rm=T)) %>% ungroup() %>% slice_max(cases, n=10) %>% pull(country)
```

#### Part b

Make a plot of smoothed lines of daily deaths over time for these 10 countries, from the beginning of the data collection to 6th July. Transform the `country` variable so that the legend is ordered in order of deaths on 6th July. The y-axis should also be log-transformed (`scale_y_log10()`). This should look something like this.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(coronavirus)
coronavirus %>% filter(date==max(date), type=='death') %>% group_by(country) %>% summarise(cases = sum(cases, na.rm=T)) %>% ungroup() %>% slice_max(cases, n=10) %>% pull(country) -> country_top10

coronavirus %>% filter(type=='death', country %in% country_top10, date <= '2020-07-06') %>% 
  ggplot(aes(x = date, y = cases, 
             color = fct_reorder2(country, date, cases)))+
  geom_smooth(se=FALSE) + 
  scale_y_log10('Number of deaths') +
  labs(x = '', color = 'Country',
       title = 'Patterns of daily deaths globally from COVID-19',
       caption = 'Johns Hopkins University CCSE, 2020')
```

