---
title: "HW 4 Solution"
author: "Abhijit Dasgupta"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "hw_img", fig.ext = "png")
library(tidyverse)
library(broom)
library(ggpubr)
datadir <- "~/Dropbox/teaching/BIOF439/data"
riskfactors <- readRDS(fs::path(datadir, "riskfactors.rds"))
```


In this homework you will generate your own R Markdown file, either using the R Markdown templates in RStudio or from scratch. For each question, you will create R chunks which you will populate with code. The code and the results should be visible in the HTML file that you knit the R Markdown to. 

The header of the R Markdown file will look like this

```
---
title: "Week 4 assignment"
author: "Your name"
output: html_document
---
```

Each question is 10 points.

### Question 1

We will use the dataset `riskfactors` available in the class data link. 

There are two files holding the same data. 
`riskfactors.rds` stores the data in R-native format, which can be read as `riskfactors <- readRDS('data/riskfactors.rds')` (you can update the path based on where you store the file). The `riskfactors.csv` file can be read using `read_csv`. 

```{r, echo=T, eval=F}
riskfactors <- readRDS("data/riskfactors.rds")
```

#### Part a

Draw a summary plot that shows data types and missing data patterns

```{r}
library(visdat)
vis_dat(riskfactors)
```


#### Part b

Draw an upset plot of missing data patterns in this dataset including the top 10 most missing variables.

```{r, echo=TRUE, eval=TRUE}
naniar::gg_miss_upset(riskfactors, nsets = 10)
```


### Question 2

Plot the frequency distribution of marital status in this dataset, as seen below.

```{r, echo=TRUE}
ggplot(riskfactors %>% filter(!is.na(marital)), aes(fct_rev(fct_infreq(marital)))) +
  geom_bar() +
  labs(
    x = "Marital status",
    y = "Frequency",
    title = "Frequency distribution of marital status"
  ) +
  coord_flip()
```

> Important points for this plot. 
> 
> 1. I do not want a bar for missing values
> 1. The tricky part may be getting the right order. See the forcats cheatsheet for hints
> 1. There is use of `coord_flip`. This should be the last thing you do in the ggplot pipe. 

```{r, echo = TRUE, eval = TRUE}
plot_data <- riskfactors %>% filter(is.na(marital))

ggplot(
  plot_data,
  aes(x = fct_rev(fct_infreq(marital)))
) +
  geom_bar() +
  labs(
    x = "Marital status",
    y = "Frequency",
    title = "Frequency distribution of marital status"
  ) +
  coord_flip()
```


### Question 3

We will use the `USArrests` data set that is inbuilt in R. 

Recreate the following plot in R

```{r, echo=TRUE}
library(ggrepel)
USArrests %>%
  rownames_to_column("State") %>%
  ggplot(aes(UrbanPop / 100, Murder)) +
  geom_text_repel(aes(label = State)) +
  labs(
    y = "Murder arrests (per 100,000)",
    caption = "McNeil (1977). Interactive Data Analysis"
  ) +
  scale_x_continuous("Percent of population that is urban",
    labels = scales::percent
  ) +
  theme_classic()
```

> You will need to add the rownames in the data to become a proper column named `State`  using `USArrests %>% rownames_to_column('State')`


### Question 4

Using the `riskfactors` data, create a panel of plots using any of the three packages introduced in lecture, where the plots will be

1. A histogram of age
1. A frequency bar plot of sex
1. A boxplot of BMI by education level
1. A barplot showing the proportion of each educational level for each marital status.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
library(ggpubr)
library(patchwork)

plt1 <- ggplot(riskfactors) +
  geom_histogram(aes(x = age))
plt2 <- ggplot(riskfactors) +
  geom_bar(aes(x = sex))
plt3 <- ggplot(riskfactors) +
  geom_boxplot(aes(x = education, y = bmi))
plt4 <- ggplot(riskfactors) +
  geom_bar(aes(x = marital, fill = education), position = "fill") +
  coord_flip()

# plot_grid(plt1, plt2, plt3, plt4, nrow=2)

# ggarrange(plt1, plt2, plt3, plt4, nrow=2, ncol=2)

(plt1 + plt2) / (plt3 + plt4)
```


### Extra credit (10 pts)

Install the **coronavirus** package (`install.packages('coronavirus')`). Once that is installed, run `coronavirus::update_dataset()`. This will update the data to the latest data from JHU, and also ask you to restart
your R session (`Session > Restart R`). 

#### Part a

The `coronavirus` dataset is in the **coronavirus** package and contains data on cases, deaths and recoveries on a daily basis from around the world. We will only use the deaths data, which is obtained by filtering the rows
of the `coronavirus` dataset to contain only rows with `type=="death"`

Extract out data for 15 February, 2021 (`date == "2021-02-15"`) and then identify the top 10 countries in terms of
deaths on that date. 


There are a couple of nuances here. For some countries, there is also provincial data, so you have to sum over those rows within country (`%>% group_by(country) %>% summarise(cases = sum(cases, na.rm=T) %>% ungroup()`))

You can use `slice_max` to grab the approriate rows. To pull out the countries, you can then use `%>% pull(country)` and save the result. You should have

```{r, echo = TRUE, message=FALSE}
library(coronavirus)
date_choice <- "2021-02-15"
coronavirus %>%
  filter(date == date_choice, type == "death") %>%
  group_by(country) %>%
  summarise(cases = sum(cases, na.rm = T)) %>%
  ungroup() %>%
  slice_max(cases, n = 10) %>%
  pull(country)
```

#### Part b

Make a plot of smoothed lines of daily deaths over time for these 10 countries, from the beginning of the data collection to 6th July. Transform the `country` variable so that the legend is ordered in order of deaths on 6th July. The y-axis should also be log-transformed (`scale_y_log10()`). This should look something like this.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(coronavirus)
library(tidyquant)
date_choice <- "2021-02-15"
coronavirus %>%
  filter(date == date_choice, type == "death") %>%
  group_by(country) %>%
  summarise(cases = sum(cases, na.rm = T)) %>%
  ungroup() %>%
  slice_max(cases, n = 10) %>%
  pull(country) -> country_top10

coronavirus %>%
  filter(type == "death", country %in% country_top10, 
         date <= date_choice) %>%
  ggplot(aes(
    x = date, y = cases,
    color = fct_reorder2(country, date, cases)
  )) +
  # geom_smooth(se=FALSE) +
  geom_ma(n = 7, linetype = 1) +
  scale_y_log10("Number of daily deaths") +
  labs(
    x = "", color = "Country",
    title = "Patterns of daily deaths globally from COVID-19 (7-day average)",
    caption = "Johns Hopkins University CCSE, 2020"
  ) +
  theme_bw()
```
