---
title: "Animated graphics in R"
author: Abhijit Dasgupta, PhD
abstract: Animated graphics in R, with mouseovers
---


```{r setup, include=FALSE}
setwd(here::here('slides/lectures'))
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)
library(palmerpenguins)
library(htmltools)
library(htmlwidgets)
library(widgetframe)
widgetdir = here::here('slides','lectures',
                       'animation2_files',
                       'figure-html',
                       'widgets')
theme_set(theme_classic()+theme(axis.text = element_text(size=14),
                                axis.title = element_text(size=16),
                                legend.text = element_text(size=14),
                                legend.title = element_text(size=16)))
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F, comment = ' ', 
                      widgetframe_self_contained = FALSE,
                      widgetframe_isolate_widgets = TRUE)
Sys.setenv("RSTUDIO_PANDOC" = '/usr/bin/')
options(htmltools.dir.version = FALSE, htmltools.preserve.raw = FALSE)
widget_show <- function(widget, filename, 
                        lib = widgetdir,...){
  saveWidget(p, file.path(widgetdir, filename), selfcontained = F, libdir = lib)
htmltools::tags$iframe(
  src = file.path(lib, filename),
  scrolling='no', frameBorder='0',width='100%', ...)
                        }
```

layout: true

<div class="my-header">
<span>BIOF 439: Data Visualization using R</span></div>

---
class: middle,inverse

# Dynamic graphics


---

## Dynamic graphics in R

There are several package in R that provide dynamic graphics meant to be consumed on the web.

Many of these are wrappers around well-known Javascript libraries like D3.js, leaflet.js and others

These packages have mostly come under the umbrella of of the `htmlwidgets` package, which allows these
HTML-based graphics to be displayed through R and R Markdown

<iframe src="https://www.htmlwidgets.org" scrolling="yes" seamless="seamless" frameBorder="1",
  width="100%", height="300"></iframe>
---

## Dynamic graphics in R

There are several broad categories of dynamic graphs

General purpose:

```{r animation-1, echo=FALSE }
tribble(
    ~Package, ~Description,
    'r2d3' , 'Interface for D3.js, requires D3.js code',
    'plotly', 'Interface with plot.ly, direct conversion from ggplot2',
    'highcharter', 'Using the Highcharts.js package',
    'dygraphs', 'For time series or longitudinal data'
) %>%
    knitr::kable() %>%
    kable_styling()

```

Maps: 

```{r animation-2 }
tribble(
    ~Package, ~Description,
        "leaflet", "Maps using OpenStreetMaps") %>%
    kable() %>%
    kable_styling()
```

---

## Dynamic graphics in R

Networks:

```{r animation-3 }
tribble(~Package, ~Description,
        'networkD3', 'Dynamic network visualizations using D3',
        'visNetwork', 'Interface to the vis.js library') %>%
    kable() %>% kable_styling()
        
```


---
class: middle, inverse

# Plotly
https://plotly.com/graphing-libraries

---

## Plotly

Plotly is a company that developed the `plotly.js` graphing library, as well as packages for R and Python. 

For the R package, it developed a turnkey method to convert `ggplot2` graphics into interactive graphs.

.pull-left[
```{r g1, echo=TRUE, fig.height=3}
plt <- ggplot(penguins,
              aes(bill_length_mm, body_mass_g,
                  color = species))+
    geom_point()
plt
```
]
.pull-right[
```{r animation-4, echo=TRUE, eval=F,fig.height=4}
ggplotly(plt)
```
```{r, echo=F}
p <- ggplotly(plt)
widget_show(p, 'animation4.html', height=300)
```

]
---

## Plotly

You can do linked plots in plotly, so interactions in one plot are reflected in a second plot. This is called _brushing_. The key here is to use `highlight_key`, which allows a data frame to be shared between multiple
plots at the same time. 

```{r brush, echo=T, eval=F,fig.height=4, fig.width=9}
d <- highlight_key(penguins)
plt1 <- ggplot(d, aes(x = bill_length_mm, y = body_mass_g))+geom_point()
plt2 <- ggplot(d, aes(x = bill_length_mm, y = flipper_length_mm))+geom_point()
subplot(plt1, plt2) %>% 
    layout(title = "Click and drag to select points") %>%
    highlight(on = "plotly_selected")
```
```{r brush_out, echo=F, eval=T,fig.height=4, fig.width=9}
d <-  highlight_key(penguins, ~species)
plt1 <- plot_ly(d, x = ~bill_length_mm, y = ~body_mass_g, mode='markers', hoverinfo = 'x+y')
plt2 <- plot_ly(d, x = ~bill_length_mm, y=~flipper_length_mm, mode='markers', hoverinfo='x+y')
p <- subplot(plt1, plt2) %>% 
    layout(title = "Click and drag to select points") %>%
    highlight(on = "plotly_selected", off='plotly_doubleclick')
widget_show(p, 'animation_brush.html', height='250')
```
