---
title: "Animated graphics in R"
author: Abhijit Dasgupta, PhD
abstract: Animated graphics in R, with mouseovers
---


```{r setup, include=FALSE}
setwd(here::here('slides/lectures'))
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)
library(palmerpenguins)
library(htmltools)
library(htmlwidgets)
library(widgetframe)
library(fs)
topdir <- here::here('slides','lectures')
widgetdir = file.path(topdir,
                       'animation2_files',
                       'figure-html',
                       'widgets')
if(!dir.exists(widgetdir)) dir.create(widgetdir)
theme_set(theme_classic()+theme(axis.text = element_text(size=14),
                                axis.title = element_text(size=16),
                                legend.text = element_text(size=14),
                                legend.title = element_text(size=16)))
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F, comment = ' ', 
                      widgetframe_self_contained = FALSE,
                      widgetframe_isolate_widgets = TRUE)
#Sys.setenv("RSTUDIO_PANDOC" = '/usr/bin/')
options(htmltools.dir.version = FALSE, htmltools.preserve.raw = FALSE)

save_tags <- function (tags, file, selfcontained = F, libdir = widgetdir) 
{
  # https://github.com/r-spatial/mapview/issues/35#issuecomment-237900713
  if (is.null(libdir)) {
    libdir <- paste(tools::file_path_sans_ext(basename(file)), 
                    "_files", sep = "")
  }
  htmltools::save_html(tags, file = file, libdir = libdir)
  if (selfcontained) {
    if (!htmlwidgets:::pandoc_available()) {
      stop("Saving a widget with selfcontained = TRUE requires pandoc. For details see:\n", 
           "https://github.com/rstudio/rmarkdown/blob/master/PANDOC.md")
    }
    htmlwidgets:::pandoc_self_contained_html(file, file)
    unlink(libdir, recursive = TRUE)
  }
  return(file)
}
widget_show <- function(widget, filename, 
                        lib = widgetdir,...){
  x <- save_tags(widget, file.path(lib, filename), selfcontained = F, libdir = lib)
return(htmltools::tags$iframe(
  src = path_rel(x, topdir),
  scrolling='no', frameBorder='0',width='100%', ...))
}

```

layout: true

<div class="my-header">
<span>BIOF 439: Data Visualization using R</span></div>

---
class: middle,inverse

# Dynamic graphics


---

## Dynamic graphics in R

There are several package in R that provide dynamic graphics meant to be consumed on the web.

Many of these are wrappers around well-known Javascript libraries like D3.js, leaflet.js and others

These packages have mostly come under the umbrella of of the `htmlwidgets` package, which allows these
HTML-based graphics to be displayed through R and R Markdown

<iframe src="https://www.htmlwidgets.org" scrolling="yes" seamless="seamless" frameBorder="1",
  width="100%", height="300"></iframe>
---

## Dynamic graphics in R

There are several broad categories of dynamic graphs

General purpose:

```{r animation-1, echo=FALSE }
tribble(
    ~Package, ~Description,
    'r2d3' , 'Interface for D3.js, requires D3.js code',
    'plotly', 'Interface with plot.ly, direct conversion from ggplot2',
    'highcharter', 'Using the Highcharts.js package',
    'dygraphs', 'For time series or longitudinal data'
) %>%
    knitr::kable() %>%
    kable_styling()

```

Maps: 

```{r animation-2 }
tribble(
    ~Package, ~Description,
        "leaflet", "Maps using OpenStreetMaps") %>%
    kable() %>%
    kable_styling()
```

---

## Dynamic graphics in R

Networks:

```{r animation-3 }
tribble(~Package, ~Description,
        'networkD3', 'Dynamic network visualizations using D3',
        'visNetwork', 'Interface to the vis.js library') %>%
    kable() %>% kable_styling()
        
```


---
class: middle, inverse

# Plotly
https://plotly.com/graphing-libraries

---

## Plotly

Plotly is a company that developed the `plotly.js` graphing library, as well as packages for R and Python. 

For the R package, it developed a turnkey method to convert `ggplot2` graphics into interactive graphs.

.pull-left[
```{r g1, echo=TRUE, fig.height=3}
plt <- ggplot(penguins,
              aes(bill_length_mm, body_mass_g,
                  color = species))+
    geom_point()
plt
```
]
.pull-right[
```{r animation-4, echo=TRUE, eval=F,fig.height=4}
ggplotly(plt)
```
```{r, echo=F}
p <- ggplotly(plt)
widget_show(p, 'animation4.html', height="100%")
```

]
---

## Plotly

You can do some customization on the tooltips (what shows up when you put your mouse over a point). 

```{r animation-5, fig.height=5, fig.width=12, echo=TRUE, eval=F}
p <- ggplot(data = diamonds, aes(x = cut, fill = clarity))+
    geom_bar(position = 'dodge')
ggplotly(p, tooltip = c('cut'))
```

```{r animation-5-out, fig.height=5, fig.width=12, echo=F, eval=T}
p <- ggplot(data = diamonds, aes(x = cut, fill = clarity))+
    geom_bar(position = 'dodge')
p1 = ggplotly(p, tooltip=c('cut'))
widget_show(p1 , 'animation5.html', height="100%") #<<

```

---

## Plotly

You can do linked plots in plotly, so interactions in one plot are reflected in a second plot. This is called _brushing_. The key here is to use `highlight_key`, which allows a data frame to be shared between multiple
plots at the same time. 

```{r brush, echo=T, eval=F,fig.height=4, fig.width=9}
penguins <- penguins %>% mutate(id = 1:n())
d <- highlight_key(penguins, ~id)
plt1 <- ggplot(d, aes(x = bill_length_mm, y = body_mass_g))+geom_point()
plt2 <- ggplot(d, aes(x = bill_length_mm, y = flipper_length_mm))+geom_point()
subplot(ggplotly(plt1), ggplotly(plt2)) %>% 
    layout(title = "Click and drag to select points") %>%
    highlight(on = "plotly_selected", off = 'plotly_doubleclick', dynamic=T)
```
```{r brush_out, echo=F, eval=T,fig.height=4, fig.width=9}
penguins <- penguins %>% mutate(id = 1:n())
d <- highlight_key(penguins, ~id)
plt1 <- ggplot(d, aes(x = bill_length_mm, y = body_mass_g))+geom_point()
plt2 <- ggplot(d, aes(x = bill_length_mm, y = flipper_length_mm))+geom_point()
p = subplot(ggplotly(plt1), ggplotly(plt2)) %>% 
    layout(title = "Click and drag to select points") %>%
    highlight(on = "plotly_selected", off = 'plotly_doubleclick', dynamic=T)
widget_show(p, 'animation_brush.html', height="100%")
```

---

## Plotly

```{r, echo=F}
widget_show(p, 'animation_brush.html', height='100%')
```
