---
title: Putting plots together
author: Abhijit Dasgupta, PhD
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F, comment = ' ', cache=TRUE,
                      fig.height=5)
setwd(here::here('slides/lectures'))
library(tidyverse)
library(palmerpenguins)
library(ggpubr)
library(cowplot)
library(patchwork)
theme_set(theme_classic()+theme(axis.text = element_text(size=14),
                                axis.title = element_text(size=16),
                                legend.text = element_text(size=14),
                                legend.title = element_text(size=16),
                                plot.title = element_text(size=18),
                                plot.subtitle = element_text(size=16),
                                plot.caption = element_text(size=12)))
```

layout: true

<div class="my-header">
<span>BIOF 439: Data Visualization using R</span></div>

---
class: middle, inverse

# Putting plots together

---

## The packages

There are three excellent packages for putting separate `ggplot` graphs together in panels.

1. **ggpubr**
1. **cowplot**
1. **patchwork**

---

## The graphs

```{r patchwork-1, eval=T, results='hide'}
plt1 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill=species)) +
  geom_boxplot()

plt2 <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g,
                             color = species))+
  geom_point()

plt3 <- ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, 
                             color = species))+
  geom_smooth(se=F)
```

```{r patchwork-2, echo=FALSE, fig.width=15, fig.height=3}
ggarrange(plt1,plt2, plt3, nrow=1)
```

---

## ggpubr

```{r patchwork-3, fig.height=6, fig.width=15, message=F}
ggarrange(plt1, plt2, plt3, ncol = 2, nrow=2)
```

---

## ggpubr

```{r patchwork-4, fig.height=6, fig.width=15, message=F}
ggarrange(plt1, plt2, plt3, ncol = 2, nrow=2, common.legend = TRUE)
```

---

## cowplot

```{r patchwork-5, fig.height=6, fig.width=15}
cowplot::plot_grid(plt1, plt2, plt3, nrow = 2, ncol = 2)
```

---

## cowplot

```{r patchwork-6, fig.height=6, fig.width=15}
cowplot::plot_grid(plt1, plt2, plt3, nrow = 2, ncol = 2, labels = c('A','B','C'))
```

---

## cowplot

```{r patchwork-7, fig.height=6, fig.width=15}
grid1 = cowplot::plot_grid(plt1, plt2, nrow = 1, rel_widths=c(0.3, 0.7),
                           labels=c('A','B'))
cowplot::plot_grid(grid1, plt3, nrow=2, labels = c('', 'C'))


```

---

## patchwork

```{r patchwork-8,fig.height=6, fig.width=15}
plt1 | plt2 / plt3
```

---

## patchwork

```{r patchwork-9,fig.height=6, fig.width=15}
(plt1 | plt2) / plt3
```

---

## patchwork

```{r patchwork-10,fig.height=6, fig.width=15}
plt1 + plt2 + plt3 + plot_layout(ncol = 2)
```


---

## Further references

ggpubr: [https://rpkgs.datanovia.com/ggpubr/](https://rpkgs.datanovia.com/ggpubr/)  
cowplot: [https://wilkelab.org/cowplot/index.html](https://wilkelab.org/cowplot/index.html) and Fundamentals of Data Visualization  
patchwork: [https://patchwork.data-imaginist.com/index.html](https://patchwork.data-imaginist.com/index.html)
