---
title: "Plotting systems in R"
author: Abhijit Dasgupta, PhD
abstract: ""
---

```{r setup, include=FALSE, message = F}
knitr::opts_chunk$set(echo = TRUE, comment = '#> ', 
                      message = F, warning = F,
                      cache=FALSE)
library(tidyverse)
library(palmerpenguins)
```

layout: true

<div class="my-header">
<span>BIOF 439: Data Visualization using R</span></div>
</div>

---

## Plotting systems in R

There are three basic plotting systems in R

+ Base plotting (what comes with R)
+ ggplot2
+ lattice

The latter two are related, in that both are based on the package **grid** that implements a system called
_grid graphics_. This actually allows ggplot2 and lattice graphics to be put together in panels using 
functions from packages like **gridExtra** and **ggpubr**

---

## ggplot2

Since we've learned **ggplot2** in this class, lets start with a plot using it

```{r plot-families-1, fig.height=5}
library(ggplot2)
library(palmerpenguins)

ggplot(penguins,
       aes(x = bill_length_mm, y = body_mass_g, color=species))+  geom_point()
```

---

## lattice

The **lattice** package was built by Deepayan Sarkar, one of the R core members

```{r plot-families-2, fig.height=5}
library(lattice)

xyplot(body_mass_g ~ bill_length_mm, groups =  species, auto.key=TRUE, data=penguins)
```

---

## Base graphics

Base graphics allow you to build up a plot, but the process is much more granular, in that you 
have to specify everything instead of a package making choices for you like **ggplot2** or **lattice**

```{r plot-families-3, fig.height=5}
plot(body_mass_g ~ bill_length_mm, data = penguins, col=species)
```

---
class: middle, inverse

# Facets

---

## Facets

Facets is where the grid-based graphics has a significant advantage over base graphics.

.pull-left[
```{r plot-families-4, fig.height=4}
ggplot(penguins,
       aes(bill_length_mm, body_mass_g))+
    geom_point()+
    facet_wrap(~species)
```
]
.pull-right[
```{r plot-families-5, fig.height=4}
xyplot(body_mass_g ~ bill_length_mm | species,
       data=penguins)
```
]

---

## Facets

```{r plot-families-6, fig.height=5, fig.width=8}
op <- par(mfrow = c(1,3))
for(n in unique(penguins$species)){
    plot(body_mass_g ~ bill_length_mm, data=subset(penguins, species==n),
         main=n, #<< Title
         ylim = range(penguins$body_mass_g, na.rm=T),
         xlim = range(penguins$bill_length_mm, na.rm=T)) #<< making scales the same
}
par(op) #<< have to return back to original graphical parameters    
```

---

## Facets

Let's consider 2-dimensional faceting

.pull-left[
```{r plot-families-7, fig.height=4}
ggplot(penguins)+
geom_point(aes(bill_length_mm,
               body_mass_g))+
    facet_wrap(species~island)
```
]
.pull-right[
```{r plot-families-8, fig.height=4}
xyplot(body_mass_g ~ bill_length_mm |
           species+island,
       data = penguins)
```
]

---

## Facets

.pull-left[
```{r base, echo=T, eval=F}
op <- par(mfrow=c(3,3))
for(i in unique(penguins$island)){
    for(j in unique(penguins$species)){
        plot(body_mass_g ~ bill_length_mm,
             data=subset(penguins, island==i & species==j),
             xlim = range(penguins$bill_length_mm,na.rm=T),
             ylim = range(penguins$body_mass_g, na.rm=T),
             main = paste(paste('Species =',j),paste('Island =',i), collapse='\n'))
    }
}
```

We basically have to do a lot of detail in base graphics
]
.pull-right[
```{r plot-families-9, eval=T, echo=F, ref.label = 'base'}

```
]

---

## Facets

+ Both ggplot2 and lattice automatically title the panels and maintain the same x- and y-axis for all theplots
- Base graphics requires a for-loop and explicitly forcing the axes to be the same. Titling is also explicitly done


---

## Why are there still 3 systems?

+ ggplot2 has traditionally been the slowest of the bunch
    + Don't try and plot 1M points
+ Base plots and lattice are a lot faster
+ Base plots and lattice aren't pipe-friendly
+ People still like base plots for the control
    + You can control almost every single aspect of a plot
    + margins, annotations and labels, aspect ratio, ...
